---
title: "Demo Class - Approximation"
subtitle: "Introduction to the Fourier Transform"
author: "Borja Cadenato"
date: "2024-04-24"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: [center, middle]
editor_options: 
  chunk_output_type: console
---

```{r libraries, echo=FALSE, message=FALSE}
library(tidyverse)
library(scales)
library(gganimate)
library(patchwork)
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)

theme_set(theme_bw())

```

```{r}
coord_freq <- function(
        xlim = c(0, 1000),
        ylim = c(0, 1.5),
        xlab = "Frequency (Hz)",
        ylab = "Amplitude") {
    
    ggplot() +
    xlim(xlim) +
    ylim(ylim) +
    geom_hline(yintercept = 0) +
    labs(
        x = xlab,
        y = ylab)
}

dirac_arrow <- function(
        frequency, 
        amplitude,
        colour = "black") {
    
    geom_segment(
        aes(
            x = frequency,
            y = 0,
            xend = frequency,
            yend = amplitude),
        lineend = "round",
        linejoin = "round",
        arrow = arrow(length = unit(0.1, "inches")),
        colour = colour)
}
```

```{r}
# An arbitrary frequency to motivate the discussion
frequency_01 <- 440

# The reference time frame for this frequency
time <- seq(0, 5 * 1 / frequency_01, length.out = 1000)

# A function to build a cosine function
get_wave <- function(
        time,
        frequency,
        amplitude) {
    
    wave = amplitude * cos(2 * pi * frequency * time)
}

# A function to represent a cosine wave over this time frame
geom_wave <- function(
        time,
        wave,
        colour = "black") {
    geom_line(
        data = tibble(
            time,
            wave),
        aes(
            x = time,
            y = wave),
        colour = colour)
}
```

class: center, middle

## How does an A4 sound wave look like?

---

class: center, middle

## An A4 sound wave

```{r}
amplitude_01 = 1

ggplot() +
    geom_wave(
        time,
        get_wave(time, frequency_01, amplitude_01)) +
    labs(x = "Time (ms)", y = "Amplitude", title = "A4 (440Hz) Cosine Wave") +
    scale_x_continuous(
        labels = label_number(scale = 1000))
```

---

class: center, top

## Representation of a wave

Frequency is measured in Hertz, number events per second.

$f_1 = 440 { \text{periods} \over \text{s}} = 440 \text{Hz}$

--

Frequency in the context of sinusoidal waves is usually measured in radians:

$\omega_1 = 2  \pi \cdot 440  {\text{radians} \over \text{s} }$

--

The representation of the 440 Hz (A4) sound wave is thus:

$x(t) = \cos(2 \pi f_1 t) = \cos(\omega_1 t)$

---

class: center, top

## An A4 sound wave in the frequency domain

```{r}
coord_freq() +
    labs(title = "A4") +
    dirac_arrow(
        frequency = frequency_01,
        amplitude = 1)
```

---

class: center, top

## Dirac Delta Function

$$
\delta(k) = 
    \begin{cases} 
    \infty & \text{if $k = 0$} \newline 
    0 & \text{if $k \neq 0$}
    \end{cases}
$$

$$
\int_{-\infty}^{\infty} \delta(k) dk = 1
$$

--

```{r fig.height=3, fig.width=3}
coord_freq(
    xlim = c(-10, 10),
    xlab = "k",
    ylab = latex2exp::TeX("$\\delta(k)$")) +
    labs(title = latex2exp::TeX("$\\delta(k)$")) +
    dirac_arrow(
        frequency = 0,
        amplitude = 1)
```

--

In general, a wave $c_1 \cdot \cos(\omega_1 t)$ of amplitude $c_1$ and angular frequency $\omega_1$ can be represented as $c_1 \cdot \delta(\omega - \omega_1)$

---

class: center, top

## An E5 sound wave

```{r echo=FALSE}
frequency_02 <- 659.2551
amplitude_02 <- 0.75

ggplot() +
    geom_wave(
        time,
        get_wave(time, frequency_02, amplitude_02)) +
    geom_wave(
        time,
        get_wave(time, frequency_01, amplitude_01),
        colour = "cornflowerblue") +
    labs(x = "Time (ms)", y = "Amplitude", title = "E5 (569.2551Hz) Cosine Wave") +
    scale_x_continuous(
        labels = label_number(scale = 1000))
```

---

class: center, middle

## An E5 sound wave in the frequency domain

```{r}
coord_freq() +
    labs(title = "E5") +
    dirac_arrow(
        frequency = frequency_02,
        amplitude = 0.75)
```

---

class: center, top

## A4 and E5 sound waves added up

```{r fig.height=3}
plot_1 <- ggplot() +
    geom_wave(
        time,
        get_wave(time, frequency_01, amplitude_01)) +
    labs(x = "Time (ms)", y = "Amplitude", title = "A4 Wave") +
    scale_x_continuous(
        labels = label_number(scale = 1000))

plot_2 <- ggplot() +
    geom_wave(
        time,
        get_wave(time, frequency_02, amplitude_02)) +
    labs(x = "Time (ms)", y = "Amplitude", title = "E5 Wave") +
    scale_x_continuous(
        labels = label_number(scale = 1000))

text_1 <- grid::textGrob(
    latex2exp::TeX("$+$"),
    gp = grid::gpar(fontsize = 32))

plot_1 + text_1 + plot_2
```

--

```{r fig.height=3, fig.width=3}
ggplot() +
    geom_wave(
        time,
        get_wave(time, frequency_01, amplitude_01) +
            get_wave(time, frequency_02, amplitude_02)) +
    labs(x = "Time (ms)", y = "Amplitude", title = "A4 + E5 Waves") +
    scale_x_continuous(
        labels = label_number(scale = 1000))
```

---

class: center, top

## A4 and E5 sound waves in the frequency domain

```{r fig.height=3}
plot_1 <- coord_freq() +
    labs(title = "A4") +
    dirac_arrow(
        frequency_01,
        1)

plot_2 <- coord_freq() +
    labs(title = "E5") +
    dirac_arrow(
        frequency_02,
        0.75)

text_1 <- grid::textGrob(
    latex2exp::TeX("$+$"),
    gp = grid::gpar(fontsize = 32))

plot_1 + text_1 + plot_2
```

--

```{r fig.height=3, fig.width=3}
coord_freq() +
    labs(title = "A4 + E5") +
    dirac_arrow(
        frequency_01,
        1) +
    dirac_arrow(
        frequency_02,
        0.75)
```

---

class: center, top

## What is a Fourier Transform?

--

If a function can be represented as a sum of sinusoidal functions:

$x(t) = c_1 \cdot \cos(\omega_1 t) + c_2 \cdot \cos(\omega_2 t) + \dots$

--

It can be represented in the frequency domain:

$X(\omega) = c_1 \cdot \delta(\omega - \omega_1) + c_2 \cdot \delta(\omega - \omega_2) + \dots$

--

We say that $X(\omega)$ is the Fourier Transform of $x(t)$

```{r fig.height=3}

plot_1 <- ggplot() +
    geom_wave(
        time,
        get_wave(time, frequency_01, amplitude_01) +
            get_wave(time, frequency_02, amplitude_02)) +
    labs(x = "Time (ms)", y = "Amplitude", title = "x(t)") +
    scale_x_continuous(
        labels = label_number(scale = 1000))

plot_2 <- coord_freq(xlab = latex2exp::TeX("\\omega")) + 
    labs(title = latex2exp::TeX("X(\\omega)")) +
    dirac_arrow(
        frequency_01,
        1) +
    dirac_arrow(
        frequency_02,
        0.75)

text_1 <- grid::textGrob(
    latex2exp::TeX("$\\leftrightarrow$"),
    gp = grid::gpar(fontsize = 32))

plot_1 + text_1 + plot_2

```

---

class: center, top

## Angular Frequency Shifting

A function $x(t) = cos(\omega t)$ can be shifted in frequency by an angular frequency offset $\omega_d$:

$x(t)_{\omega_d} = \cos([\omega - \omega_d] t)$

--

```{r fig.height=2.5, fig.width=6}
frequency_d = 120

plot_1 <- ggplot() +
    geom_wave(
        time,
        get_wave(time, frequency_01, amplitude_01)) +
    labs(
        x = "Time (ms)", 
        y = "Amplitude", 
        title = latex2exp::TeX("$\\omega_n$")) +
    scale_x_continuous(
        labels = label_number(scale = 1000))

plot_2 <- ggplot() +
    geom_wave(
        time,
        get_wave(time, frequency_01, amplitude_01),
        colour = "cornflowerblue") +
    geom_wave(
        time,
        get_wave(time, frequency_01 - frequency_d, amplitude_01)) +
    labs(
        x = "Time (ms)", 
        y = "Amplitude", 
        title = latex2exp::TeX("$\\omega_n - \\omega_d$")) +
    scale_x_continuous(
        labels = label_number(scale = 1000))

text_1 <- grid::textGrob(
    latex2exp::TeX("$\\rightarrow$"),
    gp = grid::gpar(fontsize = 32))

plot_1 + text_1 + plot_2
```

--

```{r fig.height=2.5, fig.width=6}
plot_1 <- coord_freq() +
    labs(title = latex2exp::TeX("$\\omega_n$")) +
    dirac_arrow(
        frequency_01,
        amplitude_01)

plot_2 <- coord_freq() +
    labs(title = latex2exp::TeX("$\\omega_n$")) +
    dirac_arrow(
        frequency_01,
        amplitude_01,
        colour = "cornflowerblue") +
    dirac_arrow(
        frequency_01 - frequency_d,
        amplitude_01)

text_1 <- grid::textGrob(
    latex2exp::TeX("$\\rightarrow$"),
    gp = grid::gpar(fontsize = 32))

plot_1 + text_1 + plot_2
```

---

class: center, top

## How to frequency shift a function?

Euler's formula is:

$e^{i x} = \cos x + i \sin x$

If $x = \omega t$ then

$e^{i \omega t} = \cos (\omega t) + i \sin (\omega t)$

--

Multiplying the previous expression by $e^{-i \omega_d t}$

$e^{i \omega t} \cdot e^{-i \omega_d t} = e^{i (\omega - \omega_d) t} = \cos ([\omega - \omega_d] t) + i \sin ([\omega - \omega_d] t)$

---

class: center, top

## What is the Fourier Transform of this function?

```{r}
time_02 <- seq(-0.01, 0.01, length.out = 1000)

frequency_03 <- 391.9954
amplitude_03 <- 1

frequency_04 <- 466.1638
amplitude_04 <- 0.8

frequency_05 <- 523.2511
amplitude_05 <- 0.6

ggplot() +
    geom_wave(
        time_02,
        get_wave(
            time_02, frequency_03, amplitude_03) +
        get_wave(
            time_02, frequency_04, amplitude_04) +
        get_wave(
            time_02, frequency_05, amplitude_05)) +
    labs(x = "Time (ms)", y = "Amplitude", title = "Unknown Wave") +
    scale_x_continuous(
        labels = label_number(scale = 1000))
```

---

class: center, top

## How can we identify the wave parameters?

Let's assume that the function $x(t)$ is a sum of cosine functions:

$x(t) = c_1 \cdot \cos(\omega_1 t) + c_2 \cdot \cos(\omega_2 t) + \cdots$

--

The Fourier Transform $X(\omega)$ would be

$X(\omega) = c_1 \cdot \delta(\omega - \omega_1) + c_2 \cdot \delta(\omega - \omega_2) + \cdots$

But now the parameters $c_1$, $\omega_1$, $c_2$, $\omega_2$, etc. are unknown.

---

class: center, top

## How to transform $\cos(\omega_1 t)$

Let $x(t) = \cos(\omega_1 t)$

Let's shift $\cos(\omega_1 t)$ by an angular frequency $\omega_d$

$\cos([\omega_1 - \omega_d] t)$

--

Let's call

$F(\cos(\omega_1 t), \omega) = \int_{-\infty}^{\infty} \cos([\omega_1 - \omega] t) dt$

Then, if $\omega_1 - \omega_d \neq 0$

$F(\cos(\omega_1 t), \omega_d) = \int_{-\infty}^{\infty} \cos([\omega_1 - \omega_d] t) dt = 0$

--

If $\omega_1 - \omega_d = 0$

$F(\cos(\omega_1 t), \omega_d) = \int_{-\infty}^{\infty} \cos([\omega_1 - \omega_d] t) dt = \int_{-\infty}^{\infty} dt = \infty$

because $\cos([\omega_1 - \omega_d] t) = \cos(0) = 1$

---

class: center, top

## How to transform $\cos(\omega_1 t)$

To summarise, if $x(t) = \cos(\omega_1 t)$:

$$
F(x(t), \omega) = \begin{cases}
0 & \text{if $\omega \neq \omega_1$} \newline
\infty & \text{if $\omega = \omega_1$}
\end{cases} = \delta(\omega - \omega_1)
$$

--

Let's call

$X(\omega) = F(x(t), \omega)$

Then, in general, if $x(t) = c_1 \cdot \cos(\omega_1 t)$ then

$X(\omega) = c_1 \cdot \delta(\omega - \omega_1)$

--

And if $x(t) = c_1 \cdot \cos(\omega_1 t) + c_2 \cdot \cos(\omega_2 t) + \dots$ then

$X(\omega) = c_1 \cdot \delta(\omega - \omega_1) \ c_2 \cdot \delta(\omega - \omega_2) + \dots$

---

class: center, top

## Conclusion - Fourier Transform of $x(t)$

Finally, if $x(t)$ is a general function over time, and it can be described as a (possibly infinite) series of sinusoidal functions...

--

... the Fourier Transform of $x(t)$ is a new function $X(\omega)$ in the angular frequency domain.

--

If $x(t)$ can be expressed as a series of complex exponential functions $c_k \cdot e^{i \omega_k t}$, since frequency shifting is just multiplying by $e^{- i \omega t}$, and following the same approach described before:

$$
X(\omega) = \int_{-\infty}^{\infty} x(t) \cdot e^{-i \omega t} dt
$$

---

class: center, middle

### Slides: [github.com/bcadenato/ie-demo](https://github.com/bcadenato/ie-demo)
























